Viscous Transport in Eroding Porous Media

### 2019-08-25 docs/transport/erosion_transport.tex
working on first 20 body example

In~\cite{qua-moo2018}, we observed that erosion causes the space between
nearly touching bodies to quickly expand, and flat faces develop along
the region of near contact.  This qualitative behavior is present in
Figure~\ref{fig:Eroding20vort} between bodies 3 \& 4, 15 \& 16, and
others.  However, now that we can resolve bodies that are much closer
together, we observe that, at least initially, very little erosion
occurs between certain pairs of bodies.  For instance the opening
between bodies 1 \& 6 grows much slower than the opening between bodies
15 \& 16.  A common feature of the openings that grow slowly is that
they are nearly perpendicular to the main flow direction, and this
results in very low erosion because of the small shearing.  For example,
the openings between the pair of bodies 1 \& 6, 3 \& 9, and 5 \& 13.

In this example, we also observe that erosion creates a network of
channels from the inlet to outlet where the velocity and vorticity, and
therefore erosion rate, are much larger than in other regions.  These
channels can be further visualized by considering the streamlines.  In
Figure~\ref{fig:Eroding20tracer}, we freeze the geometry at the second
time step from Figure~\ref{fig:Eroding20vort}, and we plot 200
trajectories that are initialized at $x=-1$ and equispaced in $y$.  The
trajectories are shown at 5 different times, and the final plot is a
zoom in of the lower right quadrant of the fifth time step.  There are
three clear regions where the tracers pass through the geometry fastest.
Two of these regions are located between the bodies and the solid walls
at $y=\pm 1$, and the third cuts through the porous region with the
upper part of the channel formed by bodies 1, 4, 6, and 8.  Since the
flow is fastest in these regions, the shearing is largest, and this
causes these openings to grow fastest which can be observed in
Figure~\ref{fig:Eroding20vort}.

\begin{figure}[H]
\begin{center}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b30}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b90}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b150}\\
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b210}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b270}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b270_zoom}
\caption{\label{fig:Eroding20tracer}200 trajectories in the second
geometry from Figure~\ref{fig:Eroding20vort}. The trajectories are
initialized at $x=-1$ and are equally spaced in the $y$-direction. The
first five snapshots are evenly spaced in time.  The bottom right frame
is a zoom window of the fifth snapshot.  Since we use a Barycentric
quadrature rule and the fourth-order Runge-Kutta method, the tracers can
be very close to but not penetrate the bodies.}
\end{center}
\end{figure}

Next, we use the tracer trajectories to compute the anomalous dispersion
rates and the tortuosities in the eroding geometry.  Since these
calculations use a statistical analysis of the tracer trajectories, we
increase the number of trajectories to $N_p = 1000$.  Again, the tracers
are initialized at $x=-1$, and their $y$ coordinates are distributed
evenly across the entire channel.  To compute the tortuosity, we require
the velocity at the initialization points of the tracers.  These
normalized velocities are plotted in Figure~ref{fig:Eroding20tort}(a)
for the eroded geometry with porosity 62.9\% (see
  Figure~\ref{fig:Eroding20tort}(c)).  The velocities are similar to the
  work of Matyka et al.~\cite{mat-kha-koz2008} (see Figure 4(a)), except
  that our cross-section does not cut through any of the bodies.  Next,
  in Figure~\ref{fig:Eroding20tort}(b), we plot the local
  tortuosity~\eqref{eqn:localTort} by calculating the length of each
  trajectory as it traverses the channel from $x=-1$ to $x=1$ in
  Figure~\ref{fig:Eroding20tort}(b), and dividing by the length of the
  channel.  The local tortuosity ranges from 1 to 1.27, meaning that one
  of the tracers travelled 27\% farther than it would have if the bodies
  had been absent.  The average tracer travelled 9.79\% farther than if
  the bodies had been absent, meaning that the tortuosity of the eroded
  channel is $1.098$.  Again, comparing the local tortuosity to
  Matyka~\cite{mat-kha-koz2008} (Figure 4(b)), the results are
  qualitatively similar. However, since our initial cross-section does
  not cut through any bodies, the local tortuosity does not have any
  gaps.  The discontinuities correspond to neighboring tracers with one
  traveling much farther than the other.  In
  Figure~\ref{fig:Eroding20tort}(c), we plot pairs of trajectories that
  correspond to the ten largest jumps in the local tortuosity.
  Corresponding pairs are always plotted in the same color.  The
  discrepancy between the trajectory lengths can be attributed to the
  tracers diverging from one another as they approach opposite sides of
  a stagnation point in the flow.

\begin{figure}[H]
\begin{subfigure}[b]{0.45\textwidth}
\begin{subfigure}[b]{\textwidth}
\includegraphics*[width =\linewidth]{./figs/velocity_loc20_268}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{\textwidth}
\includegraphics*[width =0.97\linewidth]{./figs/tort_local20_268}
\caption{}
\end{subfigure}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[width =\linewidth]{./figs/tort_diff_top10_268}
\caption{}
\end{subfigure}
\caption{\label{fig:Eroding20tort} The tortuosity of an eroded geometry
with porosity 62.9\%. (a) The x-component of the velocity $u(-1, y)$
normalized by its maximum velocity of $2.98 \times 10^{-3}$. (b) The
local tortuosity $\tau(y)$ on the cross section $x = -1$.  The spatially
averaged tortuosity through the channel is $1.098$.  (c) The
trajectories resulting in the ten largest differences between
neighboring trajectories.  Neighboring trajectories have the same
color.}
\end{figure}

In Figure~\ref{fig:Eroding20tort_all}, we plot the tortuosity as a
function of the porosity from the initial porosity of $0.38$ until all
the bodies have eroded.  The tortuosity is computed with both the length
of the trajectories~\eqref{eqn:tortuosity1} (red marks) and using the
spatial average of the velocity on an Eulerian
grid~\eqref{eqn:tortuosity2}.  The red square corresponds to the
porosity in Figure~\ref{fig:Eroding20tort}(c).  The methods give similar
results with any discrepancy being accounted for by slow regions of
recirculation.  As the bodies erode, this creates wide channels where
trajectories undergo only minor vertical deflections, and this explains
why the tortuosity eventually decreases with porosity.  The increase in
tortuosity for small porosities occurs because in the absence of erosion
(left plot in Figure~\ref{fig:Eroding20vort}), many of the trajectories,
such as those initialized between bodies 15 \& 18, only perform minor
deflections to pass through the narrow regions, albeit, very slowly.
However, as erosion starts to open the channels, the trajectories
deflect into the fast regions, such as the region above body 11, and
this increases the amount of vertical deflection, and therefore the
tortuosity.

\begin{figure}[H]
\center
\includegraphics*[width =0.55\linewidth]{./figs/tort_eulerian}
\caption{\label{fig:Eroding20tort_all} The tortuosity inside of an
eroding geometry initialized with 20 bodies.  The tortuosity is
calculated using the Eulerian method~\eqref{eqn:tortuosity2} (blue dots)
and Lagrangian method~\eqref{eqn:tortuosity1} (red stars).  The red
square correspond to the geometry in Figure~\ref{fig:Eroding20tort}(c).
The dash line is the fitting line $\widehat{T}(\phi)=\phi^{-p}$ as
$p=0.2064$ and the root-mean-square error is $5.9 \times 10^{-3}$.}
\end{figure}

\begin{figure}[H]
\center
\includegraphics*[width =0.55\linewidth]{./figs/20b_second_moment_long_ref}
\caption{\label{fig:Eroding20anomalousMedium} The analysis of anomalous
dispersion in 20 nearly touching bodies with respect to initial porosity
(as p=37.68\%) and another six porosities during the eroding. The dashed line
correspondes to a porosity of 100\% and the dashed-dotted lines are lines
of best fit with slope 1.0644 (as p=95.1\%), 1.0673 (as p=85.09\%), 
1.0594 (as p=75.15\%), 0.9664 (as p=65.09\%), 
0.7766 (as p=55.10\%), 0.7475 (as p=45.08\%), 
and 0.5607 (as p=37.68\%.}
\end{figure}

>>>>

In~\cite{qua-moo2018}, we observed that erosion causes the space between
nearly touching bodies to quickly expand, and flat faces develop along
the region of near contact.  This qualitative behavior is present in
Figure~\ref{fig:Eroding20vort} between bodies 3 \& 4, 15 \& 16, and
others.  However, now that we can resolve bodies that are much closer
together, we observe that, at least initially, very little erosion
occurs between certain pairs of bodies.  For instance the opening
between bodies 1 \& 6, 3 \& 9, and 5 \& 13 grow much slower than the
opening between bodies 15 \& 16.  A common feature of the openings that
grow slowly is that they are nearly perpendicular to the main flow
direction, and this results in very low erosion because of the slow flow
between the bodies.

We also observe that erosion creates a network of channels from the
inlet to outlet where the velocity and vorticity, and therefore erosion
rate, are much larger when compared to other regions.  These channels
can be further visualized by considering the streamlines.  In
Figure~\ref{fig:Eroding20tracer}, we freeze the geometry at the second
time step from Figure~\ref{fig:Eroding20vort}, and we plot 200
trajectories that are initialized at $x=-1$ and equispaced in $y$.  The
trajectories are shown at five different times, and the final plot is a
zoom in of the lower right quadrant of the fifth time step.  Since we
use a high-order quadrature rule and a high-order time stepper, we are
able to resolve trajectories that come very close to the eroded bodies.
There are three clear regions where the tracers pass through the
geometry fastest.  Two of these regions are located between the bodies
and the solid walls at $y=\pm 1$, and the third cuts through the porous
region with the upper part of the channel formed by bodies 1, 4, 6, and
8.  Since the flow is fastest in these regions, the shearing is largest,
and this causes these openings to grow fastest which can be observed in
Figure~\ref{fig:Eroding20vort}.

\begin{figure}[H]
\begin{center}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b30}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b90}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b150}\\
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b210}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b270}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b270_zoom}
\caption{\label{fig:Eroding20tracer} 200 trajectories in the second
geometry from Figure~\ref{fig:Eroding20vort}. The trajectories are
initialized at $x=-1$ and are equally spaced in the $y$-direction. The
first five snapshots are evenly spaced in time.  The bottom right frame
is a zoom in of the fifth snapshot.  Since we use the Barycentric
quadrature rule and the fourth-order Runge-Kutta method, the tracers can
be very close to but not unphysically penetrate the bodies.}
\end{center}
\end{figure}

Next, we use the tracer trajectories to compute the anomalous dispersion
rates and the tortuosities in the eroding geometry.  Since these
calculations use a statistical analysis of the tracer trajectories, we
increase the number of trajectories to $N_p =
1000$~\cite{bel-sal-rin1992}.  Again, the tracers are initialized at
$x=-1$, and their $y$ coordinates are distributed evenly across the
entire channel.  To compute the tortuosity, we require the velocity at
the initialization points of the tracers.  These normalized velocities
are plotted in Figure~ref{fig:Eroding20tort}(a) for the eroded geometry
with porosity 62.9\% (see Figure~\ref{fig:Eroding20tort}(c)).  The
velocities are similar to the work of Matyka et
al.~\cite{mat-kha-koz2008} (see Figure 4(a)), except that our
cross-section does not cut through any of the bodies.  Next, in
Figure~\ref{fig:Eroding20tort}(b), we plot the local
tortuosity~\eqref{eqn:localTort} by calculating the relative length of
each trajectory as it traverses the channel from $x=-1$ to $x=1$.  The
local tortuosity ranges from 1 to 1.27, meaning that one of the tracers
travelled 27\% farther than it would have if the bodies had been absent.
The average tracer travelled 9.79\% farther than if the bodies had been
absent, meaning that the tortuosity of the eroded channel is $1.098$.
Again, comparing the local tortuosity to Matyka~\cite{mat-kha-koz2008}
(Figure 4(b)), the results are qualitatively similar. However, since our
initial cross-section does not cut through any bodies, the local
tortuosity does not have any gaps.  The discontinuities correspond to
neighboring tracers with one traveling much farther than the other.  In
Figure~\ref{fig:Eroding20tort}(c), we plot pairs of trajectories that
correspond to the ten largest jumps in the local tortuosity.
Corresponding pairs are plotted in the same color.  The discrepancy
between the trajectory lengths can be attributed to the tracers
diverging from one another as they approach opposite sides of a
stagnation point in the flow.

\begin{figure}[H]
\begin{subfigure}[b]{0.45\textwidth}
\begin{subfigure}[b]{\textwidth}
\includegraphics*[width =\linewidth]{./figs/velocity_loc20_268}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{\textwidth}
\includegraphics*[width =0.97\linewidth]{./figs/tort_local20_268}
\caption{}
\end{subfigure}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[width =\linewidth]{./figs/tort_diff_top10_268}
\caption{}
\end{subfigure}
\caption{\label{fig:Eroding20tort} The tortuosity of an eroded geometry
with porosity 62.9\%. (a) The x-component of the velocity $u(-1, y)$
normalized by its maximum velocity of $2.98 \times 10^{-3}$. (b) The
local tortuosity $\tau(y)$ along the cross section $x = -1$.  The
spatially averaged tortuosity through the channel is $1.098$.  (c) The
trajectories resulting in the ten largest differences between
neighboring trajectories.  Neighboring trajectories have the same
color.}
\end{figure}

In Figure~\ref{fig:Eroding20tort_all}, we plot the tortuosity as a
function of the porosity from the initial porosity of $0.38$ until all
the bodies have eroded.  The tortuosity is computed with both the length
of the trajectories~\eqref{eqn:tortuosity1} (red marks) and using the
spatial average of the velocity on an Eulerian
grid~\eqref{eqn:tortuosity2}.  The red square corresponds to the
porosity in Figure~\ref{fig:Eroding20tort}(c).  The two tortuosity
calculations give similar results with any discrepancy being accounted
for by slow regions of recirculation.  As the bodies erode, this creates
wide channels where trajectories undergo only minor vertical
deflections, and this explains why the tortuosity eventually decreases
with porosity.  We computed computed lines of best fit using the
porosity-tortuosity models~\eqref{eqn:tortuosityModels}, and the power
law model gave the smallest error.  The  black
dashed line in Figure~\ref{fig:Eroding20tort_all} is the line of best
fit
\begin{align}
  \widehat{T}(\phi) = \phi^{-0.2064}
\end{align}
with a root-mean-square error of $5.9 \times 10^{-3}$.  The initial
increase in tortuosity occurs because in the absence of erosion (left
plot in Figure~\ref{fig:Eroding20vort}), many of the trajectories, such
as those initialized between bodies 15 \& 18, only perform minor
deflections to pass through the narrow regions, albeit, very slowly.
However, as erosion starts to open the channels, the trajectories
deflect into the fast regions, such as the region above body 11, and
this increases the amount of vertical deflection, and therefore the
tortuosity.

\todo[inline]{BQ: Need to fix the references to this new figure}

\begin{figure}[H]
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height = 0.8\linewidth]{./figs/tort_eulerian}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height = 0.8\linewidth]{./figs/20b_second_moment_long_ref}
\caption{}
\end{subfigure}
\caption{\label{fig:Eroding20Tranposrt} (a): The tortuosity inside of an
eroding geometry initialized with 20 bodies.  The tortuosity is
calculated using the Eulerian method~\eqref{eqn:tortuosity2} (blue dots)
and Lagrangian method~\eqref{eqn:tortuosity1} (red stars).  The red
square correspond to the geometry in Figure~\ref{fig:Eroding20tort}(c).
The dash line is the fitting line $\widehat{T}(\phi)=\phi^{-p}$ as
$p=0.2064$ and the root-mean-square error is $5.9 \times 10^{-3}$. (b):
The analysis of anomalous dispersion in 20 nearly touching bodies with
respect to initial porosity (as p=37.68\%) and another six porosities
during the eroding. The dashed line correspondes to a porosity of 100\%
and the dashed-dotted lines are lines of best fit with slope 1.0644 (as
p=95.1\%), 1.0673 (as p=85.09\%), 1.0594 (as p=75.15\%), 0.9664 (as
p=65.09\%), 0.7766 (as p=55.10\%), 0.7475 (as p=45.08\%), and 0.5607 (as
p=37.68\%.}
\end{figure}

%\begin{figure}[H]
%\center
%\includegraphics*[width = 0.55\linewidth]{./figs/tort_eulerian}
%\caption{\label{fig:Eroding20tort_all} The tortuosity inside of an
%eroding geometry initialized with 20 bodies.  The tortuosity is
%calculated using the Eulerian method~\eqref{eqn:tortuosity2} (blue dots)
%and Lagrangian method~\eqref{eqn:tortuosity1} (red stars).  The red
%square correspond to the geometry in Figure~\ref{fig:Eroding20tort}(c).
%The dash line is the fitting line $\widehat{T}(\phi)=\phi^{-p}$ as
%$p=0.2064$ and the root-mean-square error is $5.9 \times 10^{-3}$.}
%\end{figure}
%
%
%\begin{figure}[H]
%\center
%\includegraphics*[width =0.55\linewidth]{./figs/20b_second_moment_long_ref}
%\caption{\label{fig:Eroding20anomalousMedium} The analysis of anomalous
%dispersion in 20 nearly touching bodies with respect to initial porosity
%(as p=37.68\%) and another six porosities during the eroding. The dashed
%line correspondes to a porosity of 100\% and the dashed-dotted lines are
%lines of best fit with slope 1.0644 (as p=95.1\%), 1.0673 (as
%p=85.09\%), 1.0594 (as p=75.15\%), 0.9664 (as p=65.09\%), 0.7766 (as
%p=55.10\%), 0.7475 (as p=45.08\%), and 0.5607 (as p=37.68\%.}
%\end{figure}
%
